---
title: "OBDS Gene Set Enrichment and Pathway Analysis Practical"
author: "Charlie George"
date: "15/10/2021"
output: html_document
---

This is the pathway enrichment analysis practical of the OBDS course

-   The first part of this practical is using gprofiler to perform ORA on output dataset of differentially expressed genes from DESeq2

-   The major aim of this exercise is to get you used to reading online tool documentation and playing with different parameters.

# load packages

```{r setup, include=FALSE}
# load packages
library(tidyverse)
library(gprofiler2)
```

# Read in our tables from this morning in a new Rmd file

```{r}
# read in results table we output from DESeq2 this morning using tidyverse read_csv()
all_genes_df <- read_csv("CD8_vs_CD4_DE_results_lfcshrinkage_allgenes.csv") # Relative path to symlink to CSV file in project folder

# check its dimensions - and look at the top of the file -> what does this tell us about our table? 
# dim = 22,535 -> this contains all genes -> need to filter to get those tested 
all_genes_df
```

We want to perform a classic Over representation analysis (ORA) using gprofiler - to do this we first need to create our gene subsets

```{r}
# Subset 1 - All genes tested in differential expression (this is our background/control gene list) - i.e. remove padj == NA 
all_genes_tested_df <- all_genes_df %>% filter(!is.na(padj))
  
# Subset 2 - All sig upregulated genes (padj < 0.05. & log2FoldChange > 1)
upreg_genes_df <- all_genes_tested_df %>% filter(padj < 0.05 & log2FoldChange > 1)

# Subset 3 - All sig downregulated genes (padj < 0.05. & log2FoldChange < 1)
downreg_genes_df <- all_genes_tested_df %>% filter(padj < 0.05 & log2FoldChange < -1)

# Check the dimensions of each of your dataframes using dim()
dim(all_genes_tested_df)
dim(upreg_genes_df)
dim(downreg_genes_df)

# all_genes_df_filtered should be 17729
# upreg_genes_df should be 915
# downreg_genes_df should be 1204
```

```{r}
# from each of these dataframes get a vector of the ensembl ids 
all_gene_ids <- all_genes_tested_df$ENSEMBL
    
upreg_gene_ids <- upreg_genes_df$ENSEMBL
    
downreg_gene_ids <- downreg_genes_df$ENSEMBL
```

#Perform ORA using gprofiler

1.  Look at documentation - <https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html>

```{r}
# use the gost() function - what parameters should we include/change??
# we want to test the ORA of upreg_gene_ids vs our all_gene_ids

# get help for function

# write query and assign to output variable named `gost_results_obj`
gost_results_obj <- gost(
  query = upreg_gene_ids,
  organism = "mmusculus",
  correction_method = "fdr"
)

# what are our outputs  
gost_results_obj$result
```

```{r}
gost_results_obj$result %>% 
  select(term_name, term_size, precision, recall, p_value, significant) %>% 
  filter(significant) %>% 
  arrange(desc(precision), desc(recall))
```

```{r}
# plot our results using in built function in g:profiler -> which one is it? 

```

```{r}
# what does as_short_link = FALSE do in gost() function? 
# Return results as link to web interface
```

```{r}
# Filter results table for genesets with between 5-300 genes 
gost_results_obj$result %>% 
  filter(term_size >= 5, term_size <= 300)
```

2.  Try querying multiple gene lists at the same time (up and down regulated genes) using Multiple queries - see the documentation

```{r}

```

Can you plot these queries together?

```{r}

```

Output the get_version_info() to save the versions of the databases that were used for the analysis!!

```{r}
get_version_info()
```

Output the sessionInfo() so know what packages we used

```{r}
sessionInfo()
```

Now knit your report to get a lovely html file that you can share -\> this is good practise as it creates a permanent record of your plots/analysis output and also makes sure your script runs without error from start to finish!
