---
title: "Base R"
author: "Redwan Farooq"
date: '2023-02-01'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load packages

```

## Exercise 6

1.  Use `readRDS()` to load to an object called `/project/obds/shared/resources/2_r/my_day2matrix.rds`

```{r}
day2matrix <- readRDS("my_day2matrix.rds") # Relative path to symlink to RDS file in project folder
day2matrix
```

2.  Add row sums and means to this dataset

```{r}
day2matrix <- cbind(day2matrix,
                       row_sum = rowSums(day2matrix),
                       row_mean = rowMeans(day2matrix)
                       )
day2matrix
```

3.  Load the ToothGrowth datset using `data(ToothGrowth)`

```{r}
data("ToothGrowth")
```

4.  How do find more information on this dataset?

    The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice or ascorbic acid (a form of vitamin C and coded as VC)

5.  What columns of data do we have?

    -   Tooth length

    -   Supplement type (VC or OJ)

    -   Dose in mg/day

6.  What is the mean tooth length?

```{r}
mean(ToothGrowth$len)
```

7.  What is the maximum and minimum of the teeth length?

```{r}
max(ToothGrowth$len)
min(ToothGrowth$len)
```

8.  Can you calculate `rowSums` and `rowMeans` on this data?

    No, due to non-numeric values in some columns.

```{r}
summary(ToothGrowth)
```

## Exercise 7

1.  Load the AirQuality dataset.

```{r}
data("airquality")
```

2.  Look up more information about this dataset.

    Daily air quality measurements in New York, May to September 1973.\
    Daily readings of the following air quality values for May 1, 1973 (a Tuesday) to September 30, 1973.

    -   `Ozone`: Mean ozone in parts per billion from 1300 to 1500 hours at Roosevelt Island

    -   `Solar.R`: Solar radiation in Langleys in the frequency band 4000--7700 Angstroms from 0800 to 1200 hours at Central Park

    -   `Wind`: Average wind speed in miles per hour at 0700 and 1000 hours at LaGuardia Airport

    -   `Temp`: Maximum daily temperature in degrees Fahrenheit at La Guardia Airport.

3.  Examine the dataset -- how many rows and columns are there?

```{r}
airquality
```

4.  Display the column headers.

```{r}
colnames(airquality)
```

5.  Sort the dataframe by the first column (Ozone) in ascending order.

```{r}
airquality[order(airquality$Ozone), ]
```

6.  Sort the dataframe by the Month and then the Temp column in descending order and save to a new variable.

```{r}
airquality_sorted <- airquality[order(airquality$Month, airquality$Temp, decreasing = TRUE), ]
airquality_sorted
```

7.  Save this sorted data to a file on the server.

```{r}
saveRDS(airquality_sorted, file = "airquality_sorted.RDS") # Save object to RDS
write.table(
  airquality_sorted,
  file = "airquality_sorted.tsv",
  sep = "\t",
  quote = FALSE,
  row.names = FALSE,
  col.names = TRUE
) # Save values to TSV
```

## Exercise 8

1.  Create the data frames called `buildings` and `data` to merge

```{r}
buildings <- data.frame(location = c(1, 2, 3), name = c("b1", "b2", "b3"))
data <- data.frame(survey = c(1, 1, 1, 2, 2, 2), location = c(1, 2, 3, 2, 3, 1), efficiency = c(51, 64, 70, 71, 80, 58))
```

2.  What is the common key variable in these data frames?

    location

3.  Use the `merge()` function to merge the two data frames by the common key variable into a new data frame called `building_stats`.

```{r}
building_stats <- merge(buildings, data, by = "location")
building_stats
```

## Exercise 9

1.  Aggregate (or group) the `airquality` data by Month and return means on each of the numeric variables. Remove `NA` values.

```{r}
aggregate(airquality[, -5], by = list(Month = airquality$Month), FUN = mean, na.rm = TRUE)
```

2.  Aggregate the Solar.R column by Month, returning means of Solar.R. The header of column 1 should be Month. Remove `NA` values.

```{r}
aggr_solar.r_by_month <- aggregate(airquality[, "Solar.R", drop = FALSE], by = list(Month = airquality$Month), FUN = mean, na.rm = TRUE)
aggr_solar.r_by_month
```

3.  Apply the standard deviation function to the data aggregation you have just done.

```{r}
sd(aggr_solar.r_by_month$Solar.R)
```

## Exercise 10

1.  Write a function to calculate the hypotenuse of a triangle given the length of the 2 sides.

    *Remember the hypotenuse is the square root of the sum of the squares - √(a2 + b2).*

```{r}
# Calculate hypotenuse of a right angled triangle given 2 side lengths
hypotenuse <- function(a = NULL, b = NULL) {
  stopifnot(is.numeric(a), is.numeric(b)) # Check if arguments are both numeric
  out <- sqrt(a^2 + b^2)
  return(out)
}
```

2.  Run the function you have created with different values.

```{r}
hypotenuse(4, 5)
hypotenuse(1, 2)
hypotenuse(3, 7)
```

## Exercise 11

1.  Create a numeric vector of length 10. Write an `lapply` and `sapply` statement to square each element. Compare the two outputs.

```{r}
vector1 <- 1:10

l_out <- lapply(vector1, function(x) x ^ 2)
s_out <- sapply(vector1, function(x) x ^ 2)
```

```{r}
l_out
```

```{r}
s_out
```

2.  Generate a list of length 4 containing both numeric and logical vectors. Write an `lapply` or `sapply` statement to calculate the sum of the elements in each vector.

```{r}
list1 <- list(
  num1 = 1:4,
  num2 = 8:11,
  lgl1 = c(T, F, F, T),
  lgl2 = c(F, F, F, T)
)

sapply(list1, sum)
```

3.  Write an `lapply` statement to repeat each element of each vector in your list three times e.g. 1, 4, 3 would become 1, 1, 1, 4, 4, 4, 3, 3, 3. Assign the output to a new list.

```{r}
lapply(list1, rep, each = 3)
```

## Exercise 12

1.  Write a `for` loop that iterates over the numbers 1 to 7 and prints the cube of each number using `print()`.

```{r}
for (i in 1:7) {
  print(i ^ 3)
}
```

2.  Write a `for` loop that iterates over the column names of the inbuilt `iris` dataset and print each together with the number of characters in the column name in parenthesis.\
    Example output: `Sepal.Length (12)`\
    Use the functions `print()`, `paste0()` and `nchar()`. Look up what these functions do using the help feature.

```{r}
for (col in colnames(iris)) {
  # out <- paste0(col, " (", nchar(col), ")") # Using paste0
  out <- sprintf("%s (%s)", col, nchar(col)) # Using sprintf
  print(out)
}
```

3.  Write an `ifelse` loop to print the colours in `colours_vector` with four characters. Use `nchar()`.

```{r}
colours_vector <- c("red", "orange", "purple", "yellow", "pink", "blue")
four_letter_colours_vector <- ifelse(nchar(colours_vector) == 4,
                               colours_vector,
                               NA)
four_letter_colours_vector
```
