---
title: "Example code for ggplot2"
author: "Kevin Rue-Albrecht"
date: "03/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load packages
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(gridExtra)
```

# Exercise

Use `library(ggplot2)` to load the package.

```{r}
library(ggplot2)
```

## Specify the dataset.

```{r}
diamonds
```

What does `ggplot(diamonds)` do?

```{r}
ggplot(diamonds)
```

> Answer: Creates ggplot object (specify data as `diamonds` data frame)

## Add the aesthetics.

What does `ggplot(diamonds, aes(x = carat, y = price))` do?

```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

> Answer: As per `ggplot(diamonds)`, then creates x and y axes based on carat and price columns in the `diamonds` data frame

## Add geometric objects

-   Add data points showing `carat` on the x-axis and `price` on the y-axis.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.5)
```

-   Color data points by `cut`. (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut), alpha = 0.5)
```

-   Add a smoothed mean trend line. (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut), alpha = 0.2) +
  geom_smooth()
```

-   Assign that last plot to an object called `obds_diamonds`.

```{r}
obds_diamonds <- ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut), alpha = 0.2) +
  geom_smooth()
```

# Exercise

Predict the difference between these two plots

### Plot 1

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
```

### Plot 2

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut)) +
  geom_smooth()
```

> Answer: Plot 1 will create a separate (coloured) trend line for each cut, plot 2 will create a single trend line for all data points

# Exercise

## Trend lines

Using the `ChickWeight` data set:

```{r}
ChickWeight
```

-   Create a scatter plot of weight (y-axis) over time (x-axis).

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight)) +
  geom_point()
```

-   Color by diet. (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight)) +
  geom_point(aes(colour = Diet))
```

-   Add a linear mean trend line for each diet. (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_smooth(method = "lm")
```

At this point you should be able to visually identify the diet that leads to the highest mean increase in weight.

> Answer: Diet 3

-   Facet a ribbon of sub-plots, one per diet. (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_smooth(method = "lm")  +
  facet_wrap(~ Diet)
```

-   Assign that last plot to an object called `obds_chickweight`.

```{r}
obds_chickweight <- ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_smooth(method = "lm")  +
  facet_wrap(~ Diet)
```

# Exercise

## Bar plot

-   Load the `ggplot2::msleep` data set.

```{r}
msleep
```

-   Draw a bar plot of number of observations (i.e., rows) for each taxonomic order (i.e, one plot and one bar per taxonomic order).

```{r}
ggplot(msleep) +
  geom_bar(aes(x = order))
```

-   Change the angle and font size of the text for the x-axis ticks (not the axis titles). Justify the text of those x-axis ticks as right-aligned. (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep) +
  geom_bar(aes(x = order)) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

-   Change the value and font size of the title for both x and y axes. (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep) +
  geom_bar(aes(x = order)) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(size = 12)
  ) +
  labs(
    x = "Order",
    y = "Number of species"
  )
```

-   Fill each bar with colors, proportionally to the count of each genus. (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep) +
  geom_bar(aes(x = order, fill = genus)) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(size = 12)
  ) +
  labs(
    x = "Order",
    y = "Number of species"
  )
```

From this point onwards, you may need to iteratively resize the text of the ticks and axes for readability.

-   Reduce the legend key size. (Recommendation: use `unit(2, "mm")`). (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep) +
  geom_bar(aes(x = order, fill = genus)) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(size = 12),
    legend.key.size = unit(2, "mm")
  ) +
  labs(
    x = "Order",
    y = "Number of species",
  )
```

-   Force the legend to be display in 3 columns. (Recommendation: use `guide_legend(...)`). (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep) +
  geom_bar(aes(x = order, fill = genus)) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(size = 12),
    legend.key.size = unit(2, "mm")
  ) +
  labs(
    x = "Order",
    y = "Number of species"
  ) +
  guides(
    fill = guide_legend(title = "Genus", ncol = 3)
  )
```

-   Add a contour of thin black lines to the bars.

```{r}
ggplot(msleep) +
  geom_bar(aes(x = order, fill = genus), colour = "black", size = 0.1) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(size = 12),
    legend.key.size = unit(2, "mm")
  ) +
  labs(
    x = "Order",
    y = "Number of species"
  ) +
  guides(
    fill = guide_legend(title = "Genus", ncol = 3)
  )
```

-   Assign that last plot to an object called `obds_msleep`.

```{r}
obds_msleep <- ggplot(msleep) +
  geom_bar(aes(x = order, fill = genus), colour = "black", size = 0.1) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(size = 12),
    legend.key.size = unit(2, "mm")
  ) +
  labs(
    x = "Order",
    y = "Number of species"
  ) +
  guides(
    fill = guide_legend(title = "Genus", ncol = 3)
  )
```

# Exercise

## Plotting grid

-   Collate the plots that we assigned to objects through the day, as a single plot.

    -   Plots: `obds_diamonds`, `obds_chickweight`, `obds_msleep`.

    -   Methods: `cowplot::plot_grid()`, `patchwork`, `gridExtra::grid.arrange()`.

Using `cowplot`.

```{r, fig.height=8, fig.width=4}
cowplot_combined <- plot_grid(obds_diamonds, obds_chickweight, obds_msleep, ncol = 1)
cowplot_combined
```

Using `patchwork`.

```{r, fig.height=8, fig.width=4}
patchwork_combined <- obds_diamonds / obds_chickweight / obds_msleep
patchwork_combined
```

Using `gridExtra`.

```{r, fig.height=8, fig.width=4}
grid.arrange(obds_diamonds, obds_chickweight, obds_msleep)
```

-   Export the new plot in a PDF file, and open it in a PDF viewer (e.g. Adobe Acrobat Reader DC).

You will likely need a few attempts to fine-tune the width and height of the output file.

```{r}
ggsave(filename = "combined_plots.pdf", plot = cowplot_combined, width = 2000, height = 4000, units = "px")
```

# Exercise

## Pair programming

-   Explore the data set `ggplot2::mpg` and generate the most informative plot that you can!

```{r}
mpg
```

```{r, fig.height = 3, fig.width = 4}
ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point(aes(colour = year)) +
  geom_abline(slope = 1, intercept = 0) +
  facet_wrap(~ manufacturer, ncol = 5) +
  coord_equal() +
  theme(
    aspect.ratio = 0.9, 
    plot.subtitle = element_text(face = "italic")
  ) +
  labs (
    title = "Car mileage by manufacturer",
    subtitle = "from mpg dataset",
    x = "City mileage (mpg)",
    y = "Highway mileage (mpg)"
  ) +
  guides(
    colour = guide_colourbar(title = "Year")
  )
ggsave(filename = "mpg_plot.pdf", width = 2400, height = 1800, unit = "px")
```
